---
title: "Homework 1"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F)
```

```{r, results = 'hide'}
library(psych)
library(lme4)
library(knitr)
library(kableExtra)
library(qqplotr)
library(plyr)
library(tidyverse)
```


# Question 1 
Read in the HSB.csv file and save it in a dataframe called HSB_Data, excluding the variables, pracad and disclim. Verify you have done this correctly by printing the first several lines of the dataframe. 
```{r Q1}
data_path <- "https://raw.githubusercontent.com/emoriebeck/homeworks/master/homework1/HSB.csv"
HSB <- read.csv(url(data_path), stringsAsFactors = F) %>% 
  tbl_df() %>%
  select(-pracad, -disclim)

# print first few rows
head(HSB)
```

# Question 2 
Produce basic descriptive information for just these two variables: mathach and ses, using a single command. 
```{r}
describe(HSB %>% select(mathach, ses))
```

# Question 3 
What is the overall correlation between mathach and ses? 
```{r}
with(HSB, cor(mathach, ses))
```

# Question 4 
Produce a cross-classification table for female and minority. Make sure that the rows and columns of the table have appropriate labels (not just numbers).  
```{r}
HSB %>% 
  mutate(female = mapvalues(female, 0:1, c("male", "female")),
         minority = mapvalues(minority, 0:1, c("non-minority", "minority"))) %>%
  group_by(female, minority) %>%
  summarize(n = n()) %>%
  spread(key = female, value = n) %>%
  kable(., "latex", booktabs = T) %>%
  kable_styling(latex_options = c("striped","repeat_header"), full_width = F)
```

# Question 5 
Are the two variables in Question 4 independent of each other? 
```{r}
chi <- chisq.test(HSB$minority, HSB$female)
```

Yes, a $\chi^2$ test of independence suggests that minority status and gender are independent, $\chi^2$(`r chi$parameter`) = `r round(chi$statistic, 2)`, $p$ < `r round(chi$p.value, 2)`.  

# Question 6 
Produce a histogram for ses. Include a blue vertical line indicating the mean, a red vertical line indicating the median, and the normal density curve (in green). Make sure the axes are appropriately labeled. Do the data seem to be normally distributed? 
```{r}
tmp <- HSB %>%
  mutate(sdses = sd(ses, na.rm = T), medses = median(ses, na.rm = T),
         gmeanses = mean(ses, na.rm = T))
tmp2 <- tmp %>%
  gather(key = measure, value = value, gmeanses, medses)
tmp %>%
  ggplot(aes(x = ses)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.25, color = "black", 
                 fill = "lightgray") + 
  stat_function(fun = dnorm, size = 1.25, color = "darkgreen", 
                args = list(mean = HSB$gmeanses, sd = HSB$sdses)) +
  geom_vline(data = tmp2, aes(xintercept = value, color = measure)) +
  labs(x = "SES", y = "Density") +
  theme_classic() +
  theme(legend.position = "bottom")
```

# Question 7 
Produce a Q-Q plot for ses. Does this change your opinion regarding normality? 
```{r}
HSB %>% 
  ggplot(aes(sample=ses)) + 
  stat_qq_band() +
  stat_qq_line() + 
  stat_qq_point() +
  labs(x = "Actual", y = "Theoretical", 
       title = "Q-Q Plot of Studentized Residuals") +
  theme_classic()
```

# Question 8 
Produce a scatterplot of meanses (y axis) versus size (x axis). Make sure the axes are appropriately labeled. Add the best-fitting linear regression line as well as a loess (nonlinear) fit line. Color the plot symbols so that public schools are red and Catholic schools are blue. What does this figure tell you? 
```{r}
HSB %>% 
  mutate(sector = mapvalues(sector, 0:1, c("Public", "Catholic"))) %>%
  ggplot(aes(x = size, y = meanses, color = sector)) +
  scale_color_manual(values = c("blue", "red")) +
  geom_point(aes(color = sector), alpha = .5, size = .5) +
  geom_smooth(method = "lm", se = F) +
  geom_smooth(method = "loess", se = F) +
  labs(x = "School Size", y = "Mean SES", color = "Sector") +
  theme_classic() + 
  theme(axis.text = element_text(face = "bold"),
        axis.title = element_text(face = "bold", size = rel(1.2)),
        legend.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.position = "bottom")
```

# Question 9 
Plot the best-fitting line relating mathach to ses and include the 99% confidence interval around the line. 
```{r}
m <- lm(mathach ~ ses, data = HSB)

cbind(HSB, predict(m, interval = "conf", level = .99)) %>%
  ggplot(aes(x = ses, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .25, fill = "blue") +
  geom_line() +
  labs(x = "SES", y = "Fitted Values") +
  theme_classic()
```

# Question 10 
Produce a two-panel plot. In the upper panel, show the boxplots of mathach separately for each public school. In the lower panel, show the boxplots of mathach separately for each Catholic school. 
```{r, fig.height=10, fig.width=7}
orders <- HSB %>%
  group_by(School) %>%
  summarize(median = median(mathach, na.rm = T)) %>%
  arrange(median)

HSB %>%
  mutate(sector = mapvalues(sector, 0:1, c("Public", "Catholic")),
         School = factor(School, levels = orders$School)) %>%
  ggplot(aes(x = School, y = mathach, fill = School)) +
  geom_boxplot(size = .25) +
  coord_flip() +
  facet_grid(sector~., scales = "free_y") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = rel(.5)))
```

