---
title: "Homework 4"
author: "Emorie Beck"
date: \today
output: 
  pdf_document:
    toc: yes
    includes:
            in_header:
                header.tex
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# Workspace

## Packages
```{r packages, results='hide'}
library(psych)
library(lme4)
library(knitr)
library(kableExtra)
library(multcomp)
library(plyr)
library(tidyverse)
```

For this assignment, you will use the High School and Beyond data (HSB.csv) that were used for Homework 1.  

## Data
```{r data}
data_url <- "https://raw.githubusercontent.com/emoriebeck/homeworks/master/homework4/HSB(18).csv"
dat      <- read.csv(url(data_url)) %>% tbl_df
```


# Question 1
Create four dummy codes to represent the four possible combinations of the variables, minority and female: 

\begin{itemize}
\item minority girls (MG): minority = 1, female = 1  
\item minority boys (MB) minority = 1 and female = 0 
\item nonminority girls (NMG): minority = 0 and female = 1 
\item nonminority boys (NMB): minority = 0 and female = 0  
\end{itemize}

```{r}
dat <- dat %>% 
  mutate(
    MG  = ifelse(female == 1 & minority == 1, 1, 0),
    MB  = ifelse(female == 0 & minority == 1, 1, 0),
    NMG = ifelse(female == 1 & minority == 0, 1, 0),
    NMB = ifelse(female == 0 & minority == 0, 1, 0)
  )
```


# Question 2
Fit the following no-intercept model:   

## Part A
mathach ~-1 + MG + MB + NMG + NMB + (-1 + MG + MB + NMG + NMB|School). 
```{r}
source("https://raw.githubusercontent.com/emoriebeck/homeworks/master/table_fun.R")

mod2 <- lmer(mathach ~ -1 + MG + MB + NMG + NMB + (-1 + MG + MB + NMG + NMB|School), data = dat)
tab2 <- table_fun(mod2)
```


## Part B 
Construct weights for combining the fixed effect parameters from (a) that will test the main effect of Student Sex. 
```{r}
contra2b            <- matrix(c(-1,1,-1,1),nrow=1,ncol=4,byrow=TRUE)
row.names(contra2b) <- "female"

glht_mod2_b   <- glht(mod2, linfct = contra2b, alternative = "two.sided", rhs=0)
res_mod2_b    <- confint(glht_mod2_b, calpha = univariate_calpha()) 
res_mod2_b_df <- res_mod2_b$confint %>% data.frame()
```
There was a main effect of sex, $b = `r round(res_mod2_b_df$Estimate,2)`$, 95\% CI $[`r round(res_mod2_b_df$lwr,2)`, `r round(res_mod2_b_df$upr,2)`$].  

## Part C 
Construct weights for combining the fixed effect parameters from (a) that will test the main effect of Minority Status.  
```{r}
contra2c            <- matrix(c(1,1,-1,-1),nrow=1,ncol=4,byrow=TRUE)
row.names(contra2c) <- "minority"

glht_mod2_c   <- glht(mod2, linfct = contra2c, alternative = "two.sided", rhs=0)
res_mod2_c    <- confint(glht_mod2_c, calpha = univariate_calpha()) 
res_mod2_c_df <- res_mod2_c$confint %>% data.frame()
```

There was a main effect of sex, $b = `r round(res_mod2_c_df$Estimate,2)`$, 95\% CI $[`r round(res_mod2_c_df$lwr,2)`, `r round(res_mod2_c_df$upr,2)`$].  

## Part D 
Construct weights for combining the fixed effect parameters from (a) that will test the Student Sex x Minority Status interaction. 
```{r}
contra2d            <- matrix(c(1,-1,-1,1),nrow=1,ncol=4,byrow=TRUE)
row.names(contra2d) <- "minority:female"

glht_mod2_d   <- glht(mod2, linfct = contra2d, alternative = "two.sided", rhs=0)
res_mod2_d    <- confint(glht_mod2_c, calpha = univariate_calpha()) 
res_mod2_d_df <- res_mod2_d$confint %>% data.frame()
```

There was no interaction between gender and minority status, $b = `r round(res_mod2_d_df$Estimate,2)`$, 95\% CI $[`r round(res_mod2_d_df$lwr,2)`, `r round(res_mod2_d_df$upr,2)`$].  

## Part E
Construct weights for combining the fixed effect parameters from (a) that will test the Minority Status effect, but just for males. 
```{r}
contra2e            <- matrix(c(0,1,0,-1),nrow=1,ncol=4,byrow=TRUE)
row.names(contra2e) <- "Males: Minority v. Non-Minority"

glht_mod2_e   <- glht(mod2, linfct = contra2e, alternative = "two.sided", rhs=0)
res_mod2_e    <- confint(glht_mod2_e, calpha = univariate_calpha()) 
res_mod2_e_df <- res_mod2_e$confint %>% data.frame()
```

Minority and nonminority males differed, $b = `r round(res_mod2_e_df$Estimate,2)`$, 95\% CI $[`r round(res_mod2_e_df$lwr,2)`, `r round(res_mod2_e_df$upr,2)`$]. 

## Part F
Construct weights for combining the fixed effect parameters from (a) that will test the Student Sex effect, but just for minority students.  
```{r}
contra2f            <- matrix(c(1,-1,0,0),nrow=1,ncol=4,byrow=TRUE)
row.names(contra2f) <- "Minorities: Females v. Males"

glht_mod2_f   <- glht(mod2, linfct = contra2f, alternative = "two.sided", rhs=0)
res_mod2_f    <- confint(glht_mod2_f, calpha = univariate_calpha()) 
res_mod2_f_df <- res_mod2_f$confint %>% data.frame()
```

Male and female minorities differed in math achievement, $b = `r round(res_mod2_f_df$Estimate,2)`$, 95\% CI $[`r round(res_mod2_f_df$lwr,2)`, `r round(res_mod2_f_df$upr,2)`$].  

## Part G
Use the resulting weight matrix in `glht()` to obtain the significance tests. Which effects are significant?  
```{r, results = 'asis'}
contra2 <- rbind(contra2b, contra2c, contra2d, contra2e, contra2f)

glht_mod2   <- glht(mod2, linfct = contra2, alternative = "two.sided", rhs=0)
res_mod2    <- confint(glht_mod2, calpha = univariate_calpha()) 
res_mod2_df <- res_mod2$confint %>% data.frame() %>% mutate(term = rownames(.))

options(knitr.kable.NA = '')
res_mod2_df %>% 
  mutate(CI = sprintf("[%.2f, %.2f]", lwr, upr)) %>%
  select(term, Estimate, CI) %>%
  kable(., "latex", escape = F, booktabs = T, digits = 2,
        col.names = c( "Hypothesis", "b", "CI"),
        caption = "Question 2G glht Results") %>%
  kable_styling(full_width = F)
```

All the effects but the interaction are significant. 

# Question 3
Now fit this model:  

## Part A
mathach ~ 1 + minority + female + minority:female + (1 + minority + female + minority:female|School) 
```{r}
mod3 <- lmer(mathach ~ 1 + minority*female + (1 + minority * female | School), data = dat)
tab3 <- table_fun(mod3)
```

## Part B
Do the main effects and interaction tests resemble the results from the previous analysis?  
```{r, results = 'asis'}
res_mod2_df %>% 
  mutate(Estimate = sprintf("%.2f", Estimate),
         CI = sprintf("[%.2f, %.2f]", lwr, upr)) %>%
  select(term, Estimate, CI) %>%
  setNames(c("term", "b", "CI")) %>%
  filter(term %in% c("female", "minority", "minority:female")) %>%
  mutate(model = "No Intercept Model") %>%
  full_join(tab3 %>% filter(type == "Fixed Parts") %>%
              mutate(model = "Traditional Model") %>%
              select(-type)) %>%
  gather(key = est, value = value, b, CI) %>%
  unite(tmp, model, est, sep = ".") %>%
  spread(key = tmp, value = value) %>%
  kable(., "latex", escape = F, booktabs = T, digits = 2,
        col.names = c( "Term", rep(c("b", "CI"), times = 2)),
        caption = "Question 3B Results") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" " = 1, "No Intercept Model" = 2, "Traditional Model" = 2))
```


## Part C
Construct weights for combining the fixed effect parameters from (a) that will reproduce the means for the four groups (MB, MG, NMB, and NMG).  
```{r}
contra3 <- matrix(c(
  1, 1, 0, 0,
  1, 1, 1, 1,
  1, 0, 0, 0,
  1, 0, 1, 0
), nrow=4,ncol=4,byrow=TRUE)
rownames(contra3) <- c("MB", "MG", "NMB", "NMG")
```


## Part D
Use the weight matrix with `glht()`. How close are the means to those that were present in the fixed effect parameters for the model in Question 2?  
```{r}
glht_mod3   <- glht(mod3, linfct = contra3, alternative = "two.sided", rhs=0)
res_mod3    <- confint(glht_mod3, calpha = univariate_calpha()) 
res_mod3_df <- res_mod3$confint %>% data.frame()

res_mod3_df %>% 
  mutate(Estimate = sprintf("%.2f", Estimate),
         term = rownames(.),
         CI = sprintf("[%.2f, %.2f]", lwr, upr)) %>%
  select(term, Estimate, CI) %>%
  setNames(c("term", "b", "CI")) %>%
  mutate(model = "ghlt Estimates") %>%
  full_join(
    tab2 %>% filter(type == "Fixed Parts") %>%
              mutate(model = "No Intercept Model") %>%
              select(-type)
  ) %>%
  gather(key = est, value = value, b, CI) %>%
  unite(tmp, model, est, sep = ".") %>%
  spread(key = tmp, value = value) %>%
  kable(., "latex", escape = F, booktabs = T, digits = 2,
        col.names = c( "Term", rep(c("b", "CI"), times = 2)),
        caption = "Question 3D Results") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" " = 1, "glht Estimates" = 2, "No Intercept Model" = 2))
```

The means are nearly identical but have different confidence intervals.  

# Question 4
Fit the following model:   

## Part A
mathach ~ 1 + MB + NMG + NMB + (1 + MB + NMG + NMB|School). 
```{r}
mod4 <- lmer(mathach ~ 1 + MB + NMG + NMB + (1 + MB + NMG + NMB|School), data = dat)
tab4 <- table_fun(mod4)
```

## Part B
What do the fixed effect parameters mean in this analysis? 

The intercept ($\gamma_{00}$) represents the mean of the minority girls. The MB ($\gamma_{10}$) represenets the difference between minority girls and boys -- minority boys have higher math achivement than minority girls on average. The NMG term ($\gamma_{20}$) represents the difference between minority and non-minority girls -- non-minority girls have higher math achivement than minority girls on average. The NMB term ($\gamma_{30}$) represents the difference between minority girls and boys -- minority boys have higher math achivement than minority girls on average.  

## Part C 
Construct the weight matrix that will reproduce the means for the four groups (MB, MG, NMB, and NMG). 
```{r}
contra4c <- matrix(c(
   1, 1, 0, 0,
   1, 0, 0, 0,
   1, 0, 1, 0,
   1, 0, 0, 1
), nrow=4,ncol=4,byrow=TRUE)
rownames(contra4c) <- c("MB", "MG", "NMB", "NMG")
```

## Part D
Construct the weight matrix necessary to reproduce the tests of the two main effects and the interaction. 
```{r}
contra4d <- matrix(c(
    0,  1, -1, 1, # b0 + b0 + b2 - (b0 + b1 + b0 + b3) -- > b2 = b1 + b3
    0,  1, -1,-1, # b0 + b0 + b1 - (b0 + b2 + b0 + b3) -- > b1 = b2 + b3
    0, -1, -1, 1  # b0 + b0 + b3 - (b0 + b1 + b0 + b2) -- > b2 = b1 + b2
), nrow=3,ncol=4,byrow=TRUE)
rownames(contra4d) <- c("female", "minority", "minority:female")
```

##Part E 
Use the matrix in glht( ) and compare the results to those obtained in Question 2.  
```{r}
contra4 <- rbind(contra4c, contra4d)

glht_mod4   <- glht(mod4, linfct = contra4, alternative = "two.sided", rhs=0)
res_mod4    <- confint(glht_mod4, calpha = univariate_calpha()) 
res_mod4_df <- res_mod4$confint %>% data.frame() %>% mutate(term = row.names(.))

res_mod2_df %>% 
  mutate(Estimate = sprintf("%.2f", Estimate),
         CI = sprintf("[%.2f, %.2f]", lwr, upr)) %>%
  select(term, Estimate, CI) %>%
  setNames(c("term", "b", "CI")) %>%
  filter(term %in% c("female", "minority", "minority:female")) %>%
  mutate(model = "Q2 Model") %>% 
  full_join(
    tab2 %>% filter(type == "Fixed Parts") %>%
              mutate(model = "Q2 Model") %>%
              select(-type)
  ) %>%
  full_join(
    res_mod4_df %>% 
      mutate(model = "Q4 Model",
             Estimate = sprintf("%.2f", Estimate),
             CI = sprintf("[%.2f, %.2f]", lwr, upr)) %>%
      select(term, Estimate, CI, model) %>%
      setNames(c("term", "b", "CI", "model"))
  ) %>%
  gather(key = est, value = value, b, CI) %>%
  unite(tmp, model, est, sep = ".") %>%
  spread(key = tmp, value = value) %>%
  kable(., "latex", escape = F, booktabs = T, digits = 2,
        col.names = c( "Term", rep(c("b", "CI"), times = 2)),
        caption = "Question 4D Results") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" " = 1, "Q2 Model" = 2, "Q4 Model" = 2))

```


# Question 5 
Compare the fit for the three models using the `anova( )` function. If you named the fit objects: Fit\_1, Fit\_2, and Fit\_3, then use `anova(Fit_1, Fit_2, Fit_3)`. The result should not surprise you; why not?  

```{r}
anova(mod2, mod3, mod4)
```

In terms of fit statistics, these three models are identical because they are testing different linear combinations of the same data and hypotheses. 
