---
title: "Homework 11"
subtitle: "Psych 5068"
author: "Emorie Beck"
date: \today
output: 
  pdf_document:
    toc: yes
    includes:
            in_header:
                header.tex
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(knitr.kable.NA = '')
```

# Workspace

## Packages
```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(psych)
library(lme4)
library(knitr)
library(qqplotr)
library(influence.ME)
library(HLMdiag)
library(multcomp)
library(kableExtra)
library(optimx)
library(plyr)
library(tidyverse)
```

## Data
```{r data}
source("https://raw.githubusercontent.com/emoriebeck/homeworks/master/table_fun.R")
data_url <- "https://raw.githubusercontent.com/emoriebeck/homeworks/master/homework11/Safety_Binary(2).csv"
dat      <- data_url %>% read.csv %>% tbl_df 
```

Level 1:  
$\eta_{ij} = \beta_{0j} + \beta_{1j}*GMC\_Age + \beta_{2j}*sex$  

Level 2:  
$\beta_{0j} = \gamma_{00} + \gamma_{01}*crowded\_GMC_j + \gamma_{02}*economic_j + u_{0j}$  
$\beta_{1j} = \gamma_{10} + \gamma_{11}*crowded\_GMC_j + \gamma_{12}*economic_j + u_{1j}$  
$\beta_{2j} = \gamma_{20} + \gamma_{21}*crowded\_GMC_j + \gamma_{22}*economic_j$  

For this assignment you will extend these analyses to test nonlinearity and interactions. The data for this assignment are contained in the file, Safety\_Binary.csv.  

If you have problems getting some of the models to converge, you can try alternative optimizers and increase the number of iterations that are used before the algorithm gives up. The easiest way to do this is to create several alternative control lists and then substitute them into the model statement as needed:  
`cl1 <- glmerControl(optimizer = "bobyqa", optCtrl=list(maxfun=10000))`  
`cl2 <- glmerControl(optimizer = "Nelder_Mead" optCtrl=list(maxfun=10000))`  
`cl3 <- glmerControl(optimizer="optimx",optCtrl=list(method="nlminb",maxiter=10000))`  

You will need the `optimx` package for the last one.  
Switch to a different optimizer like this:  

`Safety\_Fit\_1 <- glmer(unsafe ~ 1 + Z\_Age + Z\_Age\_SQ + sex + (1 + Z\_Age + Z\_Age\_SQ + sex|street), data=Safety\_Data, binomial("logit"), control=cl1)`  

# Question 1  
In the data file, two of the variables are named age and crowded. Standardize them and name them Z\_Age and Z\_Crowded. Create a new variable that is the square of Z\_Age; name it Z\_Age\_SQ.  
```{r}
dat <- dat %>%
  mutate_at(vars(age, crowded), funs(z = as.numeric(scale(.)))) %>%
  rename(Z_Age = age_z, Z_Crowded = crowded_z) %>%
  mutate(Z_Age_SQ = Z_Age^2)
```

# Question 2  
Test a Level 1 model that contains Z\_Age, Z\_Age\_SQ, and sex. Level 2 should be unconditional (no predictors) with all residual variances estimated.  

```{r}
fit1 <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (Z_Age + Z_Age_SQ + sex | street), data = dat, family = binomial)
tab1 <- table_fun(fit1)
```


## Part A  
Is there a nonlinear relationship between age and feeling unsafe?  
Yes, there is a nonlinear relationship between Age and feeling unsafe (b = `r (tab1 %>% filter(term == "Z\\_Age\\_SQ"))$b`, OR = `r (tab1 %>% filter(term == "Z\\_Age\\_SQ"))$OR`, 95\% CI = `r (tab1 %>% filter(term == "Z\\_Age\\_SQ"))$CI`).  

## Part B  
Is sex still a significant predictor?  
Sex is still a significant preditor of feeling unsafe (b = `r (tab1 %>% filter(term == "sex"))$b`, OR = `r (tab1 %>% filter(term == "sex"))$OR`, 95\% CI = `r (tab1 %>% filter(term == "sex"))$CI`). The odds of a woman feeling unsafe is `r (tab1 %>% filter(term == "sex"))$OR` times higher than a man, on average.  

## Part C  
Can this model be simplified by eliminating any Level 2 variances? Test the following simplifications (the intercept is retained in all random effects specifications):  
### Part i  
Eliminate Z_Age_SQ from the random effects.  
```{r}
fit1i <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (Z_Age + sex | street), data = dat, family = binomial)
tab1i <- table_fun(fit1i)
```
### Part ii  
Eliminate Z_Age and Z_Age_SQ from the random effects.  
```{r}
fit1ii <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (sex | street), data = dat, family = binomial)
tab1ii <- table_fun(fit1ii)
```

### Part iii  
Eliminate sex from the random effects.  
```{r}
fit1iii <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (sex | street), data = dat, family = binomial)
tab1iii <- table_fun(fit1iii)
```

### Part iv  
Eliminate sex and Z_Age_SQ from the random effects.  
```{r}
fit1iv <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (Z_Age | street), data = dat, family = binomial)
tab1iv <- table_fun(fit1iv)
```

### Part v  
Eliminate sex, Z_Age, and Z_Age_SQ from the random effects.  
```{r}
fit1v <- glmer(unsafe ~ Z_Age + Z_Age_SQ + sex + (1 | street), data = dat, family = binomial)
tab1v <- table_fun(fit1v)
```

Compared to the original model, identify what you consider to be the simplest model
using the likelihood ratio test as well as AIC.  
```{r}
anova(fit1, fit1i)
anova(fit1, fit1ii)
anova(fit1, fit1iii)
anova(fit1, fit1iv)
anova(fit1, fit1v)
```

#Question 3  
Carry out a similar series of analyses, but substitute the sex:Z_age interaction for the nonlinear age effect.  
```{r}
fit3 <- glmer(unsafe ~ Z_Age*sex + (Z_Age * sex | street), data = dat, family = binomial)
tab3 <- table_fun(fit3)
```


## Part A  
Is there an interaction between age and sex?  
No, there is no interactionAge and sex (b = `r (tab3 %>% filter(term == "Z\\_Age:sex"))$b`, OR = `r (tab3 %>% filter(term == "Z\\_Age:sex"))$OR`, 95\% CI = `r (tab3 %>% filter(term == "Z\\_Age:sex"))$CI`). Sex does not moderate the relationship between age and feeling unsafe.   

## Part B  
What is the simplest Level 2 variance model that can be used? Test the following simplifications (the intercept is retained in all random effects specifications):  

### Part i  
Eliminate sex:Z_Age from the random effects.  
```{r}
fit3i <- glmer(unsafe ~ Z_Age*sex + (Z_Age + sex | street), data = dat, family = binomial)
tab3i <- table_fun(fit3i)
```


### Part ii  
Eliminate Z_Age and sex:Z_Age from the random effects.  
```{r}
fit3ii <- glmer(unsafe ~ Z_Age*sex + (sex | street), data = dat, family = binomial)
tab3ii <- table_fun(fit3ii)
```


### Part iii  
Eliminate sex and sex:Z_Age from the random effects.  
```{r}
fit3iii <- glmer(unsafe ~ Z_Age*sex + (Z_Age | street), data = dat, family = binomial)
tab3iii <- table_fun(fit3iii)
```

### Part iv  
Eliminate sex, Z_Age, and sex:Z_Age from the random effects.  
```{r}
fit3iv <- glmer(unsafe ~ Z_Age*sex + (1 | street), data = dat, family = binomial)
tab3iv <- table_fun(fit3iv)
```

Compared to the original model, identify what you consider to be the simplest model using the likelihood ratio test as well as AIC.  
```{r}
anova(fit3, fit3i)
anova(fit3, fit3ii)
anova(fit3, fit3iii)
anova(fit3, fit3iv)
```


# Question 4  
Now add Z\_Crowded to the simplest model from Question 3 and include all two-way interactions and the three-way interaction.  
```{r}
fit4 <- glmer(unsafe ~ Z_Age*sex + (Z_Age * sex | street), data = dat, family = binomial)
tab3 <- table_fun(fit3)
```


## Part A  
What is the highest order effect that is significant?  

## Part B  
Plot the highest order significant effect with probability of feeling unsafe as the outcome. 

## Part C  
Provide an interpretation for the plotted effect.
