---
title: "HLM (Psych 5068) Homework Assignment 8"
author: "Jessica Nicosia"
date: "Spring 2018"
output:
  pdf_document:  
    toc: yes
    toc_depth: 6
editor_options: 
  chunk_output_type: inline
---

```{r}
options(replace.assign=TRUE,width=65, digits=4,scipen=4,fig.width=4,fig.height=4)
# Clear the workspace and console.
rm(list = ls(all = TRUE)) 
cat("\014")
# Set the contrast option; important for ANOVAs.
options(contrasts = c('contr.sum','contr.poly'))
library(knitr)
```

```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(lme4)
library(arm)
library(psych)
library(lmtest)
library(car)
library(lsmeans)
library(nlme)
library(multcomp)
library(ggplot2)
library(dplyr)
library(pander)
library(ggplot2)
library(ggthemes)
library(lme4)
library(HH)
library(plot3D)
library(optimx)
library(nlme)
library(reghelper)
library(knitr)
library(lmerTest)
library(stargazer)
library(sjPlot)
library(broom)
library(plyr)
```

```{r}
# Get the data from the working directory.
setwd("/Volumes/j.nicosiaSD 1/HLM/HW8/")
```

Question 1
----------
\textbf{1. First, create a grand mean centered version of the family size variable. Name it fam size GMC.}

```{r, tidy = T}
# read in HSB data
data <- read.csv("/Volumes/j.nicosiaSD 1/HLM/HW8/iq.csv", header = TRUE)

# save in dataframe 
data <- as.data.frame(data)
```


Question 2
----------
\textbf{2. Begin wiht the following model (Model 1), which treats all coefficients as random. Is ther any evidence for birth order or family size effects in this analysis?}

```{r, tidy = T}

```

Question 3
----------
\textbf{3. Modify the first model (call the modification Model 2) in a way that will provide an omnibus test for family size effect when the two models are compared. (Hint: The models should differ by 4 degrees of freedom). What does this model comparison tell you about the presence of family size effects in this sample?}

```{r, tidy = T}

```

Question 4
----------
\textbf{4. Now modify Model 2 to remove all birth order effects (call the new model Model 3) and conduct a model comparison to Model 2.}

```{r, tidy = T}

```

Question 4a
----------
\textbf{(a) What can you conclude from this comparison?}
```{r, tidy = T}

```

Question 4b
----------
\textbf{(b) Compare Model 3 to Model 1. What is being tested and what does it offer beyond the previous two model comparisons?}
```{r, tidy = T}

```

Question 5
----------
\textbf{5. Using the most parsimonious of the three previous models, modify it (call it Model 4) to provide a model comparison that tests the significance of the true score correlation between verbal and spatial abilities at the level of the child.}
```{r, tidy = T}

```



